# Portfolio Theory with Matrix Algebra {#Portfolio-Theory-with-Matrix-Algebra}

Updated: May 13, 2021

Copyright © Eric Zivot 2015, 2016, 2021


In this chapter, we extend the mean-variance portfolio theory introduced
in chapter \@ref(Introduction-to-Portfolio-Theory) to handle
more than two risky assets. This extension allows the theory to be
applied to the analysis of real world portfolios that consist of many
risky assets as well as a risk-free asset. When working with large
portfolios, the algebra of representing portfolio expected returns
and variances becomes extremely cumbersome. As shown in chapter \@ref(Matrix-Algebra-Review),
the use of matrix algebra can greatly simplify many of these computations.
Matrix algebra formulations are also very useful when it comes to do actual computations on the computer. The matrix algebra formulas
are easy to translate into matrix programming languages like R. Popular
spreadsheet programs like Microsoft Excel, which are the workhorse
programs of many financial houses, can also handle basic matrix calculations.
All of this makes it worthwhile to become familiar with matrix techniques
for portfolio calculations.


This chapter is laid out as follows. Section 1 describes portfolios
with $N$ risky assets using matrix algebra. The concept of portfolio
risk diversification is elaborated using calculations for large portfolios.
Sections 2 through 5 cover the matrix algebra calculations required
for determining mean-variance efficient portfolios. Explicit formulas
are given for the global minimum variance portfolio, a minimum variance
portfolio that achieves a specified target expected return, and the
tangency portfolio. A simple algorithm to easily trace out the risky
asset only efficient frontier is also presented. Section 6 discusses
some practical problems associated with very large portfolios, and
section 7 introduces the **IntroCompFinR** functions for portfolio
analysis. Section 9 gives a real-world application of the theory to
asset allocation among Vanguard mutual funds.


The examples in this chapter use the **corrplot**, **IntroCompFinR**
and **PerformanceAnalytics** packages. Make sure these packages
are installed and loaded in R before replicating the chapter examples.

```{r, echo=TRUE}
suppressPackageStartupMessages(library(corrplot))
suppressPackageStartupMessages(library(IntroCompFinR))
suppressPackageStartupMessages(library(PerformanceAnalytics))
options(digits=3)
```

## Portfolio Analysis Functions in R {#Portfolio-Analysis-Functions}
\index{Portfolio analysis functions}
The package **IntroCompFinR** contains a few R functions for
computing Markowitz mean-variance efficient portfolios allowing for
short sales using matrix algebra computations. These functions allow
for the easy computation of the global minimum variance portfolio,
an efficient portfolio with a given target expected return, the tangency
portfolio, and the efficient frontier. These functions are summarized
in Table \@ref(tab:TablePorfolioFunctions).


------------------------------------------------------------------------------------------
**Function**                **Description**
------------------------    --------------------------------------------------------------
`getPortfolio`              create portfolio object 
  
`globalMin.portfolio`       compute global minimum variance portfolio 
  
`efficient.portfolio`       compute minimum variance portfolio subject to target return 
  
`tangency.portfolio`        compute tangency portfolio 
  
`efficient.frontier`        compute efficient frontier of risky assets 
------------------------------------------------------------------------------------------

Table: (\#tab:TablePorfolioFunctions) **IntroCompFinR** functions for computing mean-variance efficient portfolios


The following examples illustrate the use of the functions in Table
\@ref(tab:TablePorfolioFunctions) using the example data in Table
\@ref(tab:Table-ThreeAssetExample):

```{r echo=TRUE}
asset.names <- c("MSFT", "NORD", "SBUX")
mu.vec = c(0.0427, 0.0015, 0.0285)
sigma.mat = matrix(c(0.0100, 0.0018, 0.0011,
           0.0018, 0.0109, 0.0026,
           0.0011, 0.0026, 0.0199),
         nrow=3, ncol=3)
r.f=0.005
mu.vec
sigma.mat
r.f
```

To specify a `portfolio` object, you need an expected
return vector and covariance matrix for the assets under consideration
as well as a vector of portfolio weights. To create an equally weighted
portfolio use:

```{r echo=TRUE}

library(IntroCompFinR)
ew = rep(1,3)/3
equalWeight.portfolio = getPortfolio(er=mu.vec,cov.mat=sigma.mat,weights=ew)
class(equalWeight.portfolio)
```

Portfolio objects have the following components:

```{r echo=TRUE}

names(equalWeight.portfolio)
```

There are `print()`, `summary()` and `plot()`
methods for portfolio objects. The `print()` method gives:

```{r echo=TRUE}

equalWeight.portfolio
```

The `plot()` method shows a bar chart of the portfolio weights:

```{r PortMatrixPlotMethod, fig.cap='Plot method for objects of class "portfolio"', out.width='80%', fig.asp=.75, fig.align='center', echo=TRUE}

plot(equalWeight.portfolio, col="cornflowerblue")
```


The resulting plot is shown in Figure \@ref(fig:PortMatrixPlotMethod).
The global minimum variance portfolio (allowing for short sales) $\mathbf{m}$
solves the optimization problem \@ref(eq:PortMatrixGlobalMinObjective)
and is computed using \@ref(eq:PortMatrixGlobalMinSolution). To
compute this portfolio use the function `globalMin.portfolio()`:

```{r echo=TRUE}

gmin.port = globalMin.portfolio(mu.vec, sigma.mat)
class(gmin.port)
gmin.port
```

A mean-variance efficient portfolio $\mathbf{x}$ that achieves the
target expected return $\mu_{0}$ solves the optimization problem
\@ref(eq:PortMatrixEfficientPortfolio2) and is computed using \@ref(eq:PortMatrixEfficientPortfolioSolution).
To compute this portfolio for the target expected return $\mu_{0}=E[R_{\textrm{msft}}]=0.0427$
use the `efficient.portfolio()` function:

```{r echo=TRUE}

target.return = mu.vec[1]
e.port.msft = efficient.portfolio(mu.vec, sigma.mat, target.return)
class(e.port.msft)
e.port.msft
```

The tangency portfolio $\mathbf{t}$ is the portfolio of risky assets
with the highest Sharpe's slope and has solutions given by \@ref(eq:PortMatrixTangencySolution).
To compute this portfolio with $r_{f}=0.005$ use the `tangency.portfolio()`
function:

```{r echo=TRUE}

tan.port = tangency.portfolio(mu.vec, sigma.mat, r.f)
class(tan.port)
tan.port
```

The the set of efficient portfolios of risky assets can be computed
as a convex combination of any two efficient portfolios. It is convenient
to use the global minimum variance portfolio as one portfolio and
an efficient portfolio with target expected return equal to the maximum
expected return of the assets under consideration as the other portfolio.
Call these portfolios $\mathbf{m}$ and $\mathbf{x}$, respectively.
For any number $\alpha$, another efficient portfolio can be computed
as: 
\[
\mathbf{z}=\alpha\mathbf{x}+(1-\alpha)\mathbf{m}
\]
The function `efficient.frontier()` constructs the set of efficient
portfolios using this method for a collection of $\alpha$ values
on an equally spaced grid between $\alpha_{min}$ and $\alpha_{max}$.
For example, to compute 20 efficient portfolios for values of $\alpha$
between -2 and 1.5 use:

```{r echo=TRUE}

ef = efficient.frontier(mu.vec, sigma.mat, alpha.min=-0.5, 
                        alpha.max=2, nport=20)
attributes(ef)
ef
```

Use the `summary()` method to show the weights of these portfolios.
Use the `plot()` method to create a simple plot the efficient
frontier:

```{r PortMatrixMarkowitzPlotMethod, fig.cap='Plot method for objects of class "Markowitz"', out.width='80%', fig.asp=.75, fig.align='center', echo=TRUE}

plot(ef)
```


The resulting plot is shown in Figure \@ref(fig:PortMatrixMarkowitzPlotMethod).
To create a more elaborate plot of the efficient frontier showing
the original assets and the tangency portfolio use:

```{r PortMatrixEfficientFrontierPlot, fig.cap='Efficient frontier plot.', out.width='80%', fig.asp=.75, fig.align='center', echo=TRUE}

plot(ef, plot.assets=T, col="blue", pch=16) 
points(gmin.port$sd, gmin.port$er, col="green", pch=16, cex=2) 
text(gmin.port$sd, gmin.port$er, labels = "Global min", pos = 4) 
points(tan.port$sd, tan.port$er, col="red", pch=16, cex=2) 
text(tan.port$sd, tan.port$er, labels = "Tangency", pos = 3) 
sr.tan = (tan.port$er - r.f)/tan.port$sd 
abline(a=r.f, b=sr.tan, col="green", lwd=2)
```

The resulting plot is shown in Figure \@ref(fig:PortMatrixEfficientFrontierPlot).

